---
title: "Using ppidyom with humdrumR"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using ppidyom with humdrumR}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = TRUE, cache = TRUE, cache.path = 'cache/')
library(ppidyom)
library(data.table)
```




Though it can be used on any R data vectors, `ppidyom` is fully incorporated into [humdrumR](https://humdrumr.ccml.gtcmt.gatech.edu/), making it very easy to apply to existing musical datasets using various operational definitions ("viewpoints").

```{r, message = FALSE}
library(humdrumR)
library(ppidyom)

```


# First examples

Let's start with trusty Bach chorales, the first ten of which are packaged with humdrumR.
To load them, we just need this:

```{r}

chorales <- readHumdrum(humdrumRroot, "HumdrumData/BachChorales/.*krn")

```

If you have any other humdrum data on your machine, you can just pass `readHumdrum()` the path and pattern to match your files.

Before applying ppidyom, consider what dimension of the music you want to model, and your *operational definition* for that dimension.
This choral data is encoded as `**kern`, which means it has relative rhythm data and absolute pitch encoded for each note, as well as notation details like stem direction.
Here we'll consider just focus on modeling pitch, ignoring rhythm.
But "pitch" can be viewed different ways: absolute vs relative, tonal vs atonal, etc.

Let's start with simple (no octave information) scale degree, which we can extract using `solfa(simple = TRUE)`.
We can then pass that straight to `ppidyom()`:

```{r}

chorales |> solfa(simple = TRUE) |> ppidyom()


```

And that's all it takes!
We see a unique information theory prediction for each and every note in the score.

----

The ppidyom() function has *many* arguments, but here we're just using the default settings.
What exactly did ppidyom do here, by default?:

+ *Leave-one-out cross validation*:
	+ For each chorale, the model is trained on the *other nine* chorales to create the "long-term" model.
+ All four voices (bass, tenor, alto, soprano) within each chorale are used, each treated like a separate sequence.
+ The maximum N-gram length of five is used.

## Different operational definitions 


Perhaps you decide that, *given your research questions*, it makes more sense to incorporate absolute pitch information.
Just get rid of `simple = TRUE` (or set `simple = FALSE`).

```{r}

chorales |> solfa(simple = FALSE) |> ppidyom()

```

Or maybe you just want to atonal absolute pitch (semitones).
That's easy, just switch to the `semits()` function:


```{r}

chorales |> semits() |> ppidyom()

```


I wonder how different the ppm information content estimates will be when we use these different approaches.
Let's compare:


```{r}

chorales |> solfa(simple = TRUE)  |> ppidyom() |> pull() -> tonal
chorales |> semits() |> ppidyom() |> pull() -> atonal

draw(tonal, atonal)

```


# Other examples


How about a different type of pitch data: harmonies?
Let's pull up `humdrumR`'s prepackaged subset of the *Rolling Stone 200* dataset:

```{r}

rs <- readHumdrum(humdrumRroot, '/HumdrumData/RollingStoneCorpus/.*hum')

```

We can extract the roman numerals from the `**harm` spine, and then apply ppidyom:

```{r}

rs[[, '**harm']] |> ppidyom()

```

Now we get a information-content estimate for each every chord symbol.
Notice that this corpus includes separate annotations from two different analyses (Temperley and de Clercq), and ppidyom() is using both as the training data.
We might want to just use on or the other.




