---
title: "N-grams 101"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{N-grams 101}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
library(dplyr)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = TRUE)
```


Music, like language, consists of streams of sequential events.
For example, a melody consists of a sequence of notes, while a chord progression consists of sequences of chords.
"N-grams" are one approach to analyzing sequential data like this. 

It is probably intuitive that each note in a melody is perceived and understood in the context of the notes around it---particularly the ones before it.
But, how much context do we consider? And how do we use that context to "understand" each note?

The basic idea of N-grams is to look at a fixed number ($N$) of notes.
We might pick $N=5$, which means we'll look at windows ("grams") of five notes at a time.
We then create these grams for every *overlapping* five-note sequence in the melody---the grams becomes like a sliding 5-note window on the melody.

Consider this familiar tune, the Ode to Joy:

```{r, echo = FALSE}

ode <- c('E', 'E', 'F', 'G', 'G', 'F', 'E', 'D', 'C', 'C', 'D', 'E', 'E', 'D', 'D',
		  	 'E', 'E', 'F',' G', 'G', 'F', 'E', 'D','C','C','D', 'E','D', 'C', 'C')


cat(ode, sep = ' ')
```

This melody consists of `r length(ode)` notes.
If we divide it into five grams, we get:

```{r, echo = FALSE}

fiveode <- paste(ode, lead(ode, 1), lead(ode, 2), lead(ode, 3), lead(ode, 4))


cat(fiveode, sep = '\n')


```


Note, as we said above, that the N-grams overlap!


## Predictive context


In the example above, we break the Ode to Joy into overlapping 5-grams.
What are these grams useful for?
One particularly common approach is in predictive modeling.

In the context of music cognition (and psycholinguistic) research, n-grams are often used to model how a listener will perceive each note in a melody (or chord in a progression, etc.).
Consider a five-gram `G-F-E-D-C`.
Somebody listening to this melody in real time would hear the first four notes (`G-F-E-D-`) *then* hit the final note (`C`).
So we could think of the first four ($N-1$) notes of the gram as the context for the *last* note.
But then we look at the next 5-gram, which represents four notes of context plus one *new* note---and we can keep doing this.
With this perspective, we can try to predict the *last* note the of the N-gram using the $N-1$ notes of context---and do this over and over again for each new note!

This approach to sequential modeling is the core of "N-gram" analysis.
Generally, we compute *conditional probabilities* for consequent events, *given* $N-1$ events of context.


## Padding

You probably noticed that for the first four notes of the Ode To Joy, we can't construct predictive 5-grams, because there haven't been five notes yet.
This is a basic issue you have to deal with, and there are various approaches.
The simplest approaches are to (1) just ignore these early grams or (2) pad them with "NA" (or something like that), but there are more sophisticated approaches.

## Partial Predictive Matching


A key problem with N-gram modeling is knowing what size of N-gram to use.
You might think "the bigger the better, right"?
That *is* sort of right...but if you use large $N$, every melodic sequence becomes unique, so there is nothing to learn or *generalize*.
The core idea of the *partial predictive matching* algorithm is to compute N-grams of a range of sizes, then make your final prediction based on a blend of these different N-grams.


